New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-17 jso.qmd
squintability
fixed Definition 6, Proposition 1, and its proof.

\begin{equation}
\varsigma(f) = -4 \times \max_{d} g'(d) \times \arg \max_{d} g'(d)
\label{eq-squintability}
\end{equation}

where $4$ is a constant scaling factor, $-\max_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \max_{d} g'(d)$ represents the projection distance at which this largest gradient is attained.

@@

\begin{prop}\label{prop:convex-concave}
Let $g_1(d)$ be a convex decreasing function and $g_2(d)$ be a concave decreasing function both defined on [0, 1] where $d \in [0, D]$. Let $d_1 := \underset{d}{\arg \max} |g_1'(d)|$ and $d_2 := \underset{d}{\arg \max} |g_2'(d)|$.

$$\varsigma(g_1) < \varsigma(g_2)$$

For a convex function, $g_1'(d) < 0 \; \forall d$ and is non-decreasing. For a concave function, $g_2'(d) < 0 \; \forall d$ and is non-increasing. As such, it follows that $d_1 < d_2$.

Suppose both functions achieve the same maximum gradient $-g_{\max}$, i.e., $\max g'_1(d_1) = \max g'_2(d_2) = -g_{\max}$,

$$
\varsigma(g_1(d_1)) = -4  \max g'_1(d_1)  d_1 = -4  (-g_{\max}) d_1 \leq -4  (-g_{\max}) d_2 = -4 \max g'_2(d_2)  d_2 = \varsigma(g_2(d_2))
$$

\end{prop}

>>>>

\begin{equation}
\varsigma(f) = -4 \times \min_{d} g'(d) \times \arg \min_{d} g'(d)
\label{eq-squintability}
\end{equation}

where $4$ is a constant scaling factor, $-\min_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \min_{d} g'(d)$ represents the projection distance at which this largest gradient is attained.

@@

\begin{prop}\label{prop:convex-concave}
Let $g_1$ be a convex, strictly decreasing function and $g_2$ be a concave, strictly decreasing function, both defined on $[0, D]$ where $D>0$, and taking values in $[0, 1]$. Then
$$\varsigma(g_1) < \varsigma(g_2)$$
\end{prop}

\begin{proof}
Let $d_1 := \underset{d}{\arg \min }\,g_1'(d)$ and $d_2 := \underset{d}{\arg \min }\,g_2'(d)$.
Since $g_1$ is convex, $g_1''(d) >= 0 \; \forall d \in [0, D]$, meaning that $g_1'$ is an increasing function, taking its minimum for $d_1=0$. As a result, $\varsigma(g_1)=-4 \times g'(0) \times 0 =0$.
Since $g_2$ is concave, $g_2''(d) <= 0 \; \forall d \in [0, D]$, meaning that $g_2'$ is a decreasing function, taking its minimum for $d_2=D>0$. Since $g_2$ is strictly decreasing, $g_2'(d) < 0 \; \forall d \in [0, D]$. In particular $g_2'(D)<0$. As a result $\varsigma(g_2)=-4 \times g'(D) \times D >0$. This proves that $\varsigma(g_1) < \varsigma(g_2)$.
\end{proof}

### 2024-09-17 jso.tex
squintability
fixed Definition 6, Proposition 1, and its proof.

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criterion function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optimum has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, boxrule=0pt, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

\begin{tcolorbox}[enhanced jigsaw, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colbacktitle=quarto-callout-note-color!10!white, left=2mm, titlerule=0mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm, leftrule=.75mm, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, arc=.35mm, opacitybacktitle=0.6, breakable, toprule=.15mm, coltitle=black]

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criterion function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optimum has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, leftrule=.75mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, rightrule=.15mm, opacityback=0, breakable, coltitle=black, colframe=quarto-callout-note-color-frame, left=2mm, toprule=.15mm, bottomrule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, opacitybacktitle=0.6, colback=white]
